NanoMaLy

### 2015-11-19 paper/pldi16-submission/interactive.tex
down to 11pp

A trace, on its own, is likely too detailed to be
a good explanation of the type error. One approach
is to use the witness input to step through the
program with a \emph{debugger} to observe how
the program evolves.
%
This route is problematic for two reasons.
%
First, existing debuggers and interpreters for
typed languages (\eg\ \ocaml) typically require
a type-correct program as input.
%
Second, we wish to have a quicker way to get
to the essence of the error, \eg\ by skipping
over irrelevant sub-computations, and focusing
on the important ones.

Next, we present a novel way to debug executions.
%
First, we develop a notion of a
\emph{reduction graph}~(\S~\ref{sec:reduction-graph}).
%
Second, we extend our semantics with a form of \emph{tracing}
so that they incrementally collect the edges in the graph
(\S~\ref{sec:inter-semant}).
%
Finally, we express a set of common \emph{interactive debugging}
steps as graph traversals (\S~\ref{sec:traversing-graph}), yielding
an novel visual interactive debugger that allows the user
to effectively visualize \emph{how} the program goes (wrong.)

@@

\subsection{Reduction Graphs}

\subsection{Tracing Semantics}
\label{sec:inter-semant}
%
\begin{figure*}[t]
\relDescription{Trace Syntax}
$$
\begin{array}{rrcl}
  & \tr & ::= & \bullet \spmid \singlestep{e}{e}; \tr \spmid \subterm{e}{e}; \tr
\end{array}
$$
\\
\relDescription{\subtermssym}
\begin{gather*}
\begin{array}{lcl}
\subtermssym                 & \dcolon & e \to \tr \\
\subterms{\eapp{e_1}{e_2}}   & \defeq & \subterm{\eapp{e_1}{e_2}}{e_1}; \subterm{\eapp{e_1}{e_2}}{e_2} \\
\subterms{\eplus{e_1}{e_2}}   & \defeq & \subterm{\eplus{e_1}{e_2}}{e_1}; \subterm{\eplus{e_1}{e_2}}{e_2} \\
\subterms{\eif{e_1}{e_2}{e_3}}   & \defeq & \subterm{\eif{e_1}{e_2}{e_3}}{e_1}; \\
                                &        & \subterm{\eif{e_1}{e_2}{e_3}}{e_2}; \\
                                &        & \subterm{\eif{e_1}{e_2}{e_3}}{e_3} \\
\subterms{\elet{x}{e_1}{e_2}}   & \defeq & \subterm{\elet{x}{e_1}{e_2}}{e_1}; \\
                                &        & \subterm{\elet{x}{e_1}{e_2}}{e_2} \\
\subterms{\efun{x}{e}}       & \defeq & \subterm{\efun{x}{e}}{e} \\
\subterms{e}                 & \defeq & \bullet
\end{array}
\end{gather*}
\judgementHead{Evaluation}{\stepg{e}{\su}{\tr}{e}{\su}{\tr}}
\begin{gather*}
\inference[\recontext]
  {\stepg{e}{\su}{\tr}{e_1}{\su_1}{\tr_1}}
  {\stepg{C[e]}{\su}{\tr}{C[e_1]}{\su_1}{\singlestep{C[e]}{C[e_1]}; \subterms{C[e_1]}; \tr_1}}
\\ \\
\inference[\reappgood]
  {\pair{\efun{x}{e}}{\su_2} = \force{v_1}{\tfun{\thole{}}{\thole{}}}{\su_1}}
  {\stepg{\eapp{v_1}{v_2}}{\su_1}{\tr}
         {e\sub{x}{v_2}}{\su_1;\su_2}{\singlestep{\eapp{v_1}{v_2}}{e\sub{x}{v_2}}; \subterms{e\sub{x}{v_2}}; \tr}}
\\ \\
\inference[\reappbad]
  {\pair{\stuck}{\su_2} = \force{v_1}{\tfun{\thole{}}{\thole{}}}{\su_1}}
  {\stepg{\eapp{v_1}{v_2}}{\su_1}{\tr}{\stuck}{\su_1;\su_2}{\singlestep{\eapp{v_1}{v_2}}{\stuck}; \tr}}
\end{gather*}
\caption{A selection of the operational semantics from
  Figure~\ref{fig:operational}, extended to collect a full reduction
  graph.}
\label{fig:interactive}
\end{figure*}
%
The changes to the operational semantics of \S~\ref{sec:semantics} are
mechanical, so we will not reproduce them in full; instead we describe
the procedure for extending a transition rule and provide a selection of
examples in Figure~\ref{fig:interactive}.

First, we extend the transition relation to collect a set of edges from
which we will construct the graph.
%
An edge between two expressions is either a ``steps-to'' edge -- written
\singlestep{e_1}{e_2} -- indicating that $e_1$ transitions to $e_2$ in a
single step, or a ``sub-term'' edge -- written \subterm{e_1}{e_2} --
indicating that $e_1$ contains $e_2$ as a sub-expression.
%
Collecting the steps-to edges is a simple matter of recording the
consequent of each original rule in the trace; each original judgment
\step{e_1}{\su_1}{e_2}{\su_2} becomes
\stepg{e_1}{\su_1}{\tr_1}{e_2}{\su_2}{\singlestep{e_1}{e_2}; \tr_1}.
%
The sub-term edges can be delegated to a new \subtermssym helper
function, which adds edges from an expression to each of its
\emph{immediate} sub-expressions.
%
We collect new sub-term edges after each transition, thus the final
template for the small-step relation is:
\[
\stepg{e_1}{\su_1}{\tr_1}{e_2}{\su_2}{\singlestep{e_1}{e_2}; \subterms{e_2}; \tr_1}
\]

After evaluation the reduction graph can be constructed directly from
the trace $\tr$ as follows:
\[
G(\tr) = \pair{\{e \spmid e \in \tr\}}{\tr}
\]

>>>>

A trace, on its own, is too detailed to be
a good explanation of the type error. One approach
is to use the witness input to step through the
program with a \emph{debugger} to observe how
the program evolves.
%
This route is problematic for two reasons.
%
First, existing debuggers and interpreters for
typed languages (\eg\ \ocaml) typically require
a type-correct program as input.
%
Second, we wish to have a quicker way to get
to the essence of the error, \eg\ by skipping
over irrelevant sub-computations, and focusing
on the important ones.

Next, we present a novel way to debug executions.
%
First, we develop a notion of a \emph{reduction graphs}
and extend our semantics with a form of \emph{tracing}
so that they incrementally collect the edges
in the graph (\S~\ref{sec:inter-semant}).
%
Next, we express a set of common \emph{interactive debugging}
steps as graph traversals (\S~\ref{sec:traversing-graph}), yielding
an novel interactive debugger that allows the user
to effectively visualize \emph{how} the program goes (wrong.)

@@

\subsection{Tracing Semantics}
\label{sec:inter-semant}
%
\begin{figure*}[t]
% \relDescription{Trace Syntax}
% $$
% \begin{array}{rrcl}
  % & \tr & ::= & \bullet \spmid \singlestep{e}{e}; \tr \spmid \subterm{e}{e}; \tr
% \end{array}
% $$
% \\
\relDescription{Computing Sub-Terms}
\begin{gather*}
\begin{array}{lcl}
\subtermssym                 & \dcolon & e \to \tr \\
\subterms{\eapp{e_1}{e_2}}   & \defeq & \subterm{\eapp{e_1}{e_2}}{e_1}; \subterm{\eapp{e_1}{e_2}}{e_2} \\
\subterms{\eplus{e_1}{e_2}}   & \defeq & \subterm{\eplus{e_1}{e_2}}{e_1}; \subterm{\eplus{e_1}{e_2}}{e_2} \\
\subterms{\eif{e_1}{e_2}{e_3}}   & \defeq & \subterm{\eif{e_1}{e_2}{e_3}}{e_1}; \\
                                &        & \subterm{\eif{e_1}{e_2}{e_3}}{e_2}; \\
                                &        & \subterm{\eif{e_1}{e_2}{e_3}}{e_3} \\
\subterms{\elet{x}{e_1}{e_2}}   & \defeq & \subterm{\elet{x}{e_1}{e_2}}{e_1}; \\
                                &        & \subterm{\elet{x}{e_1}{e_2}}{e_2} \\
\subterms{\efun{x}{e}}       & \defeq & \subterm{\efun{x}{e}}{e} \\
\subterms{e}                 & \defeq & \bullet
\end{array}
\end{gather*}
\judgementHead{Traced Evaluation}{\stepg{e}{\su}{\tr}{e}{\su}{\tr}}
\begin{gather*}
\inference[\recontext]
  {\stepg{e}{\su}{\tr}{e_1}{\su_1}{\tr_1}}
  {\stepg{C[e]}{\su}{\tr}{C[e_1]}{\su_1}{\singlestep{C[e]}{C[e_1]}; \subterms{C[e_1]}; \tr_1}}
\\ \\
\inference[\reappgood]
  {\pair{\efun{x}{e}}{\su_2} = \force{v_1}{\tfun{\thole{}}{\thole{}}}{\su_1}}
  {\stepg{\eapp{v_1}{v_2}}{\su_1}{\tr}
         {e\sub{x}{v_2}}{\su_1;\su_2}{\singlestep{\eapp{v_1}{v_2}}{e\sub{x}{v_2}}; \subterms{e\sub{x}{v_2}}; \tr}}
\\ \\
\inference[\reappbad]
  {\pair{\stuck}{\su_2} = \force{v_1}{\tfun{\thole{}}{\thole{}}}{\su_1}}
  {\stepg{\eapp{v_1}{v_2}}{\su_1}{\tr}{\stuck}{\su_1;\su_2}{\singlestep{\eapp{v_1}{v_2}}{\stuck}; \tr}}
\end{gather*}
\caption{A selection of the operational semantics from
  Figure~\ref{fig:operational}, extended to collect a full reduction
  graph.}
\label{fig:interactive}
\end{figure*}

\paragraph{Reduction Graphs}
%
A \emph{steps-to} edge is a pair of expressions \singlestep{e_1}{e_2}, which
intuitively indicates that $e_1$ steps, in a single step, to $e_2$.
%
A \emph{sub-term} edge is a pair of expressions \subterm{e_1}{e_2}, which
intuitively indicates that $e_1$ contains $e_2$ as a sub-expression.
%
A \emph{reduction graph} is a set of edges:
$$\tr ::= \bullet \spmid \singlestep{e}{e}; \tr \spmid \subterm{e}{e}; \tr$$

\paragraph{Tracing Semantics}
%
We extend the transition relation (\S~\ref{sec:semantics}) to
collect the set of edges corresponding to the reduction graph.
%
Concretely, we extend the operational semantics to
a relation of the form $\stepg{e}{\vsu}{\tr}{e'}{\vsu'}{\tr'}$
where $\tr'$ collects the edges of the transition.

\paragraph{Collecting Edges}
%
Next, we describe the general recipe for extending a transition
rule to collect edges, and provide a selection of examples
in Figure~\ref{fig:interactive}.
%
The steps-to edges are collected by recording the consequent of
each original rule in the trace. That is, each original judgment
\step{e}{\vsu}{e'}{\vsu'} becomes
\stepg{e}{\vsu}{\tr}{e'}{\vsu'}{\singlestep{e}{e'}; \tr}.
%
The sub-term edges are delegated to a helper function \subtermssym\
which adds edges from an expression to each of its
\emph{immediate} sub-expressions.
%
We collect \subtermssym edges after each transition,
to get the following template for the small-step relation:
\[
\stepg{e}{\vsu}{\tr}{e'}{\vsu'}{\singlestep{e}{e'}; \subterms{e'}; \tr}
\]
\RJ{shouldn't this be $\subterms{e}$?}

% After evaluation the reduction graph can be constructed directly from
% the trace $\tr$ as follows:
% \[
% G(\tr) = \pair{\{e \spmid e \in \tr\}}{\tr}
% \]

### 2015-11-19 paper/pldi16-submission/semantics.tex
down to 11pp

\paragraph{Narrowing Types} The procedure $\force{e}{t}{\vsu}$ takes as
input an term \texttt{e} and refines it to have the type \texttt{t} by
yielding a pair of either the same expression and a value substitution
the refines the holes in \texttt{e} to ensure it has type \texttt{t}, or
yields the stuck term if no such refinement is possible. In the case
where \texttt{e} is a hole, it first checks in the given $\vsu$ to see
if the hole has already been instantiated and, if so, returns the
existing instantiation:
%
$$
\begin{array}{lcl}
% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
% \forcesym            & ::     & (e, t) \rightarrow \pair{e}{\vsu} \\
\force{\ehole{i}}{t}{\vsu} & \defeq & \hspace{-1ex}
\begin{cases}
  \pair{\vsu(i)}{\emptysu}        & \mbox{if}\ i \in \vsu \\
  \pair{v}{\ehole{i} \mapsto v}   & \mbox{o.w.} v = \gen{t}\\
\end{cases} \\
 % \begin{array}{l}
    % \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \pair{\vsu(i)}{\emptysu}\ \mathtt{else} \\
         % \elet{v}{\gen{t}}{\pair{v}{\ehole{i} \mapsto v}}
  % \end{array} \\
\force{v}{\ehole{}}{\vsu}  & \defeq & \pair{v}{\emptysu} \\
\force{n}{\tint}{\vsu}     & \defeq & \pair{n}{\emptysu} \\
\force{v}{\tint}{\vsu}     & \defeq & \pair{\stuck}{\emptysu} \\
\force{b}{\tbool}{\vsu}    & \defeq & \pair{b}{\emptysu} \\
\force{v}{\tbool}{\vsu}    & \defeq & \pair{\stuck}{\emptysu} \\
\force{\efun{x}{e}}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\efun{x}{e}}{\emptysu} \\
\force{v}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\stuck}{\emptysu} \\
\end{array}
$$

@@

Recall that a key technical challenge in generating witnesses is
that we have no (static) type information to rely upon. Thus, we
must avoid the trap of generating \emph{spurious} witnesses that
arise from picking irrelevant values, when instead there exist
perfectly good values of a \emph{different} type, under which
the program would not go wrong.

@@

\paragraph{Generating Witnesses}
% Procedure \texttt{genWitness}}
%
Finally, Figure~\ref{fig:algo-gen-witness} summarizes the overall
implementation of our search for witnesses, in the procedure @genWitness@,
which takes as input a bound @n@ and a list of top level binders, and
returns a list of witness expressions $e$ that demonstrate how the input
program gets stuck.
%
The search proceeds as follows.
%
@(1)@ we invoke @genArgs@ to determine the number of parameters required
by the (curried) target function.
%
@(2)@ we obtain a closed expression $e$ by closing the target with the
computed holes.
%
@(3)@ we take the first @n@ traces returned by @eval@ on the closed $e$
and
@(4)@ we extract the substitutions corresponding to the $\stuck$ traces,
and use them to return the list of witnesses.
%
We obtain the following as a corrollary of Theorem~\ref{thm:soundness}:

>>>>

\paragraph{Narrowing Types} The procedure $\force{e}{t}{\vsu}$ takes as
input an term \texttt{e} and refines it to have the type \texttt{t} by
yielding a pair of either the same expression and a value substitution
the refines the holes in \texttt{e} to ensure it has type \texttt{t}, or
yields the stuck term if no such refinement is possible. In the case
where \texttt{e} is a hole, it first checks in the given $\vsu$ to see
if the hole has already been instantiated and, if so, returns the
existing instantiation:
%
$$
\begin{array}{lcl}
% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
% \forcesym            & ::     & (e, t) \rightarrow \pair{e}{\vsu} \\
\force{\ehole{i}}{t}{\vsu} & \defeq & \hspace{-1ex}
\begin{cases}
  \pair{\vsu(i)}{\emptysu}        & \mbox{if}\ i \in \vsu \\
  \pair{v}{\ehole{i} \mapsto v}   & v = \gen{t}\\
\end{cases} \\
 % \begin{array}{l}
    % \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \pair{\vsu(i)}{\emptysu}\ \mathtt{else} \\
         % \elet{v}{\gen{t}}{\pair{v}{\ehole{i} \mapsto v}}
  % \end{array} \\
\force{v}{\ehole{}}{\vsu}  & \defeq & \pair{v}{\emptysu} \\
\force{n}{\tint}{\vsu}     & \defeq & \pair{n}{\emptysu} \\
\force{v}{\tint}{\vsu}     & \defeq & \pair{\stuck}{\emptysu} \\
\force{b}{\tbool}{\vsu}    & \defeq & \pair{b}{\emptysu} \\
\force{v}{\tbool}{\vsu}    & \defeq & \pair{\stuck}{\emptysu} \\
\force{\efun{x}{e}}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\efun{x}{e}}{\emptysu} \\
\force{v}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\stuck}{\emptysu} \\
\end{array}
$$

@@

A key technical challenge in generating witnesses is
that we have no (static) type information to rely upon. Thus, we
must avoid the trap of generating \emph{spurious} witnesses that
arise from picking irrelevant values, when instead there exist
perfectly good values of a \emph{different} type, under which
the program would not go wrong.

@@

\paragraph{Generating Witnesses}
% Procedure \texttt{genWitness}}
%
Finally, Figure~\ref{fig:algo-gen-witness} summarizes the overall
implementation of our search for witnesses, in the procedure @genWitness@,
which takes as input a bound @n@ and a list of top level binders, and
returns a list of witness expressions $e$ that demonstrate how the input
program gets stuck.
%
The search proceeds as follows.
%
@(1)@ we invoke @genArgs@ to determine the number of parameters required
by the (curried) target function.
%
@(2)@ we obtain a closed expression $e$ by closing the target with the
computed holes.
%
@(3)@ we take the first @n@ traces returned by @eval@ on the closed $e$
and
@(4)@ we extract the substitutions corresponding to the $\stuck$ traces,
and use them to return the list of witnesses.
%
We obtain the following corollary of Theorem~\ref{thm:soundness}:
